---
title: Using .GRP symbol with R data.table package
output:
  html_document:
    code_folding: hide
author: David Ross
date: '2021-05-12'
slug: using-grp-symbol-with-r-data-table-package
categories: []
tags:
  - R
---

I am a big fan of the [data.table](https://rdatatable.gitlab.io/data.table/) package in R for data manipulation. I use the csv reader function `fread` all the time and enjoy the concise syntax for many basic queries of the data. There are some [special symbols](https://rdatatable.gitlab.io/data.table/reference/special-symbols.html) associated with the package that I've used before, especially `.N` and `.SD`. However, I had never used the `.GRP` symbol until I faced a particular query for a class assignment recently. I am documenting how I used it in this post.

The data come from the [pnwflights14](https://github.com/ismayc/pnwflights14) package which is modeled upon the [nycflights13](https://github.com/tidyverse/nycflights13) package.

```{r}
library(data.table)
data("flights", package = "pnwflights14")
setDT(flights)
# just analyze flights originating from Portland
flights <- flights[origin == "PDX"]
```

The analysis question is: by month, what new routes were added or removed? Only the `dest` and `month` columns are necessary to answer this query. The special symbol `.GRP` is used with the data.table syntax as an index for the current group. So, I grouped the data by month and the `.GRP` symbol now aligns with the month of the current group which allows for comparison with the previous month.

```{r}
# flights dropped from previous month
dropped <- flights[order(month), .(dropped = setdiff(flights[month == .GRP - 1, dest], dest)), by = month]
# flights added from previous month
added <- flights[order(month), .(added = setdiff(dest, flights[month == .GRP - 1, dest])), by = month][month != 1]
```

```{r, echo=FALSE}
dropped <- dropped[, .(dropped = paste(c(.SD$dropped), collapse = ", ")), by=month]
added <- added[, .(added = paste(c(.SD$added), collapse = ", ")), by=month]
routes <- data.table(month = 2:12)
routes <- merge(routes, dropped, on = "month", all = TRUE)
routes <- merge(routes, added, on = "month", all = TRUE)
routes[, month := month.abb[month]]
```

I used the excellent [gt](https://gt.rstudio.com/) package for creating a formatted table.

```{r, warning=FALSE, echo=FALSE}
library(magrittr)
library(gt)
routes %>% 
 gt() %>% 
  tab_header(
    title = md("**Routes added and dropped**"),
    subtitle = "Changes in destinations from previous month"
  ) %>% 
  tab_source_note(
    source_note = "Source: pnwflights R package"
  ) %>% 
  tab_footnote(
    footnote = "For flights departing PDX in the year 2014.",
    locations = cells_title("subtitle")
  ) %>% 
  fmt_missing(
    columns = 2:3
  ) %>% 
  cols_label(
    month = "",
    dropped = "Dropped",
    added = "Added"
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "lightcyan"),
      "font-variant: small-caps;"
    ),
    locations = cells_body(columns = "month")
  )
```