---
title: UpSet plot to visualize TCR repertoire overlap
author: David Ross
date: '2021-06-11'
slug: upset-plot-to-visualize-tcr-repertoire-overlap
categories: []
tags:
  - R
  - dataviz
---
For our final project in data visualization class I explored something I'd thought about when I first came across various papers comparing T-cell receptor repertoire overlap. This challenge comes up in various contexts but for the visualization I create below it is to view the concordance between different methods of sequencing the TCR repertoire. The authors of the [methods paper](https://bmcbiotechnol.biomedcentral.com/articles/10.1186/s12896-017-0379-9) use a series of Venn diagrams with four sets to visualize intersecting sets—where the same clonotypes were detected. For me Venn diagrams break down with greater than three sets.

```{r, echo=FALSE}
# source R script to create dataset
source("Ross-final.R")
```

```{r}  
library(grid)
upset(dt, order.by = "freq", set_size.show = FALSE)
grid.text("Overlap between the most abundant TCR sequences\ndetected by each method",
          x = 0.65, y=0.95, gp=gpar(fontsize=10))
```

The UpSet plot above is based on the data presented in Figure 3a of [Rosati et al](https://bmcbiotechnol.biomedcentral.com/articles/10.1186/s12896-017-0379-9) as a Venn diagram. A slide deck for an in-class presentation on this visualization is available [here](/UpSet-presentation.pdf).